Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-04 paper/sec-mva.tex
final touches on the PU section

It is also interesting to show the distributions of kinematic variables
that according to the MVA provide a small amount of discrimination
between signal and background, and where the should not be an issue
of redundancy.
%
In Fig.~\ref{fig:mva_substructure_2}
we show the kinematic variable that carries less weight in the MVA,
    the azimuthal angle separation between the two Higgs candidates
    $\Delta\phi_{hh}$, for the resolved and boosted categories:
    according to Fig.~\ref{fig:nnweights}, essentially no discrimination power
    should be obtained from this variable, and indeed the shapes of the two signal
    and background
    are very similar here.
    %
    From Fig.~\ref{fig:nnweights} we also saw that the
    invariant mass distributions $m_h$ carry little information
    to discriminate signal over background:
    this was already verified from 
the corresponding
distributions in Fig.~\ref{fig:mHHinv}.
%
As explained before, the reduced discrimination power arises
from 
the  smearing of jet four-momenta introduced to mimic
realistically detector resolution.\footnote{We have verified that
  if no smearing is applied, the $m_h$ distributions for the leading
  and subleading Higgs candidates become the most
important variables in the MVA.}

>>>>

It is also interesting to show the distributions of kinematic variables
that according to the MVA provide a small amount of discrimination
between signal and background, and where the should not be an issue
of redundancy.
%
In Fig.~\ref{fig:mva_substructure_2}
we show the kinematic variable that carries less weight in the MVA,
    the azimuthal angle separation between the two Higgs candidates
    $\Delta\phi_{hh}$, for the resolved and boosted categories:
    according to Fig.~\ref{fig:nnweights}, essentially no discrimination power
    should be obtained from this variable, and indeed the shapes of the two signal
    and background
    are very similar here.
    %
    From Fig.~\ref{fig:nnweights} we also saw that the
    invariant mass distributions $m_h$ carry little information
    to discriminate signal over background:
    this was already verified from 
the corresponding
distributions in Fig.~\ref{fig:mHHinv}.
%
As explained before, the reduced discrimination power arises
from 
the  smearing of jet four-momenta introduced to mimic
realistically detector resolution.

### 2015-12-04 paper/sec-pileup.tex
final touches on the PU section

\clearpage
\section{Impact of pile-up}

@@

Let us now discuss in more detail the results of the MVA
classification, by showing various estimators as function
of the cut $y_{\rm cut}$ in the ANN discriminant.
%
In Fig.~\ref{fig:nev2_PU}
we show the number of signal and background events that
are expected at the HL-LHC as a function of
$y_{\rm cut}$ -
the corresponding results without PU are those of
Fig.~\ref{fig:nev2}.
%
As we can see from Fig.~\ref{fig:nev2_PU}, the trends are
similar as those in the case without PU.
%
Comparing Figs.~\ref{fig:nev2} and~\ref{fig:nev2_PU}, we also observe
that the intermediate category is substantially degraded once PU effects
are accounted for, the reason being a reduction in the number of signal
events that are now assigned to this category.
%
This shows that the definition of the intermediate category is seriously
affected by PU, as opposed to the resolved and boosted selections.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.65\textwidth]{plots/nev2_SKPU80.pdf}
\caption{\small Same as Fig.~\ref{fig:nev2} now
  with  $\la n_{\rm PU}\ra=80$ PU events 
  and SK subtraction.
}
\label{fig:nev2_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In Fig.~\ref{fig:sb_mva_PU} we show the signal significance,
$S/\sqrt{B}$, as well as the signal over background ratio,
$S/B$, as a function of the ANN discriminant $y_{\rm cut}$, in the case
of the analysis including the effects of PU
with $\la n_{\rm PU}\ra=80$.
%
The corresponding results in the case without PU were shown in
Fig.~\ref{fig:sb_mva}.
%
As can be seen, the MVA-driven enhancement is robust in the
presence of PU.
%
While unsurprisingly there us a degradation as compared to the case wo PU,
we still manage to achieve a signal significance of
around $S/\sqrt{B}\simeq 2.5$ separately for the boosted and resolved
categories.
%
On the other hand, the intermediate category becomes irrelevant.
%
We conclude that the qualitative conclusions drawn in
Sect.~\ref{sec:mva} are robust in the presence
of realistic PU effects.
%
Since no specific effort has been performed to
optimize PU subtraction, for example by tuning the value
of the patch length $a$ in {\tt SoftKiller}, we expect that
there is
still room for improvement in this respect.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.48\textwidth]{plots/ssb_SKPU80.pdf}
\includegraphics[width=0.48\textwidth]{plots/sb_SKPU80.pdf}
\caption{\small Same as Fig.~\ref{fig:sb_mva}
  now
  with  $\la n_{\rm PU}\ra=80$ PU events 
  and SK subtraction.
}
\label{fig:sb_mva_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

It is illustrative to quantify which of the input variables
to the MVA are most significant in the case of PU,
and compare it with the corresponding
results without PU.
%
We use the same estimator as in Sect.~\ref{sec:signalsignificance},
namely the sum
of the absolute value of all the weights connected to a given
input neuron $i$, Eq.~(\ref{eq:totweight}).
%
The results for the resolved and boosted categories are shown
on Fig.~\ref{fig:nnweights_PU}.
%
As we can see by comparing to the corresponding
results without PU in Fig.~\ref{fig:nnweights}, we see that
in the boosted category, the discrimination power of the invariant
mass of Higgs candidates is decreased and that of the various substructure
variables, in particular $C_2^{(\beta)}$ and
$D^{(\beta)}$, is conversely
increased.
%
This reflects the fact that these substructure variables are
relatively robust against PU contamination.

In the case of the resolved category, in the case without PU the highest
values of the weight sum were found for the $p_T$ of the leading
Higgs and for the Higgs invariant masses.
%
This is also true in the case with PU, but now the rapidity difference
between the two Higgs candidates, $\Delta y_{hh}$ increases its
importance, again reflecting that this variable is relatively
insensitive to PU effects.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.49\textwidth]{plots/res_wgthist_SKPU80.pdf}
\includegraphics[width=0.49\textwidth]{plots/bst_wgthist_SKPU80.pdf}
\vspace{-0.5cm}
\caption{\small
Same as Fig.~\ref{fig:nnweights} in the case of PU.
}
\label{fig:nnweights_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Finally, we now combine the three categories, resolved,
intermediate and boosted, and determine the signal
significance taking into account all event topologies.
%
Adding in quadrature the signal significance from the three
(exclusive) categories for the corresponding
optimal value of $y_{\rm cut}$, our final combined result is
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0 \, ,
\ee
that is, well
above the threshold for evidence, and  relatively close to the
requirement for discovery.
%

>>>>

\section{Impact of pile-up}

@@

Let us now discuss in more detail the results of the MVA
classification, by showing various estimators as function
of the cut $y_{\rm cut}$ in the ANN discriminant.
%
In Fig.~\ref{fig:nev2_PU}
we show the number of signal and background events that
are expected at the HL-LHC as a function of
$y_{\rm cut}$ -
the corresponding results without PU are those of
Fig.~\ref{fig:nev2}.
%
As we can see, the general trends are
similar as those without PU.
%
Comparing Figs.~\ref{fig:nev2} and~\ref{fig:nev2_PU}, we also observe
that the intermediate category is substantially degraded once PU effects
are accounted for, the reason being  sizable
a reduction in the number of signal
events that are now assigned to this category.
%
This shows that the selection criteria
of the intermediate category are less
resilient against PU contamination,
as opposed to the resolved and boosted selections.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.65\textwidth]{plots/nev2_SKPU80.pdf}
\caption{\small Same as Fig.~\ref{fig:nev2} in the
case of events with PU, for
 $\la n_{\rm PU}\ra=80$ 
  and SK subtraction.
}
\label{fig:nev2_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In Fig.~\ref{fig:sb_mva_PU} we show the signal significance,
$S/\sqrt{B}$, as well as the signal over background ratio,
$S/B$, including now the effects of PU.
%
The corresponding results in the case without PU were shown in
Fig.~\ref{fig:sb_mva}.
%
As can be seen, the MVA-driven enhancement is robust in the
presence of PU.
%
While unsurprisingly there us a degradation as compared to the case wo PU,
we still manage to achieve a signal significance of
around $S/\sqrt{B}\simeq 2.5$ separately for the boosted and resolved
categories.
%
On the other hand, the intermediate category becomes irrelevant.
%
We conclude that the qualitative conclusions drawn in
Sect.~\ref{sec:mva} are robust in the presence
of realistic PU effects.
%
Since no specific effort has been performed to
optimize PU subtraction, for example by tuning the value
of the patch length $a$ in {\tt SoftKiller}, we believe that
there is
still room for improvement in this respect.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.48\textwidth]{plots/ssb_SKPU80.pdf}
\includegraphics[width=0.48\textwidth]{plots/sb_SKPU80.pdf}
\caption{\small 
Same as Fig.~\ref{fig:sb_mva} in the
case of events with PU, for
 $\la n_{\rm PU}\ra=80$ 
  and SK subtraction.
}
\label{fig:sb_mva_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

It is important to quantify which of the input variables
to the MVA carry the highest discrimination power
in the case of PU,
and compare it with the corresponding
results without PU.
%
We use the same estimator as in Sect.~\ref{sec:signalsignificance},
namely the sum
of the absolute value of all the weights connected to a given
input neuron $i$, Eq.~(\ref{eq:totweight}).
%
The results for the resolved and boosted categories are shown
on Fig.~\ref{fig:nnweights_PU}.
%
As we can see by comparing to the corresponding
results without PU in Fig.~\ref{fig:nnweights}, 
in the boosted category, the discrimination power of the invariant
mass of Higgs candidates is decreased and that of the various substructure
variables, in particular $C_2^{(\beta)}$ and
$D^{(\beta)}$, is conversely
increased.
%
This reflects the fact that these substructure variables are
relatively robust against PU contamination.
%
In addition, also the $p_T$ distributions of the AK03
subjets within the large-$R$
jets carry important information.
%
In the case of the resolved category,  the highest
values of the ANN weights without PU
were found for the $p_T$ of the leading
Higgs and for the Higgs invariant masses.
%
This is also true in the case with PU, but now the rapidity difference
between the two Higgs candidates, $\Delta y_{hh}$ increases its
importance, again reflecting that this variable is relatively
insensitive to PU effects.
%

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.49\textwidth]{plots/res_wgthist_SKPU80.pdf}
\includegraphics[width=0.49\textwidth]{plots/bst_wgthist_SKPU80.pdf}
\vspace{-0.5cm}
\caption{\small
Same as Fig.~\ref{fig:nnweights} in the
case of events with PU, for
 $\la n_{\rm PU}\ra=80$ 
  and SK subtraction.
}
\label{fig:nnweights_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

At this point, we can finally combine the three categories, resolved,
intermediate and boosted, and determine the overall signal
significance taking into account all event topologies.
%
Adding in quadrature the signal significance from the three
(exclusive) categories for the corresponding
optimal value of $y_{\rm cut}$, our final combined result, accounting
for PU effects, is
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0 \, ,
\ee
to be compared to the value $\lp S/\sqrt{B}\rp_{\rm tot} \simeq 8.5$
that was obtained in the case without PU.
%
We conclude that, even when realistic PU effects are accounted
for, at the HL-LHC the measurement of
Higgs pair production in the $b\bar{b}b\bar{b}$ final state should be 
well the threshold for evidence, and even meeting the
requirement for discovery might be within reach.
%
